The state of knowledge for marine invertebrate biodiversity in the continental US

### 2018-04-16 manuscript.Rmd
incorporate GP's comments in abstract and intro

---
title: "The state of knowledge for marine invertebrate biodiversity in the continental US"
date: ' ' #'`r format(Sys.time())`'
output:
  bookdown::pdf_document2:
    includes:
    #  before_body: ../template/doc_prefix.tex
      in_header: preamble.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc: no
fontsize: 12pt
link-citations: yes
csl: nature.csl
bibliography: /home/francois/Library/dark-plankton.bib
abstract: Museum collections have been recording our knowledge about species and their geographic ranges for about 250 years, but until recently, this information was fragmented across institutions and difficult to synthesize. The digitization of biodiversity information is now enabling the synthesis of this information, and can be used to highlight knowledge gaps. Here we use publicly accessible biodiversity data to assemble a list of `r format(n_distinct(combined_species$worms_valid_name), big.mark = ",")` species of marine invertebrates from the continental United States. Our knowledge about species richness and their geographic ranges is limited by sampling effort, as `r prop_10_specimens_species`% of the species are known from less than 10 occurrences. While recent developments in high-throughput sequencing promise to greatly speed up the documentation of marine biodiversity, anchoring results from such studies into existing DNA sequence data and associated vouchers would provide a richer biological context to interpret the insights provided by these datasets. However, with `r format_output(prop_spp_barcoded$p_barcoded * 100)`% of the marine invertebrate fauna that has sequences available for the most widely used DNA barcoding gene (cytochrome oxidase I, COI), and given that for many species these records have been sampled in biogeographic areas outside the USA, additional efforts are needed to assemble a comprehensive library of reference DNA barcoding sequences. More sampling, coordinated at the national scale, is needed to improve our knowledge of species and their geographic ranges. This data will allow the detection of the impacts of climate change and other anthropogenic strains on marine biodiversity. Advances in computing, information systems, and sequencing technologies, have set the stage to inform where future efforts for documenting biodiversity should be applied.
---

@@

With increasing impact of human activities on the biosphere, we need efficient and accurate approaches to document changes in species ranges and community composition. <!-- add why? -->  While terrestrial ecosystems have received considerable attention, the baseline of healthy and resilient marine ecosystems is not well understood [@Worm2006], and modeling the impact of human activities has been limited to a few taxonomic groups [@Tittensor2010]. For marine organisms, because of the diversity of their complex life history strategies, their body size range, and the difficulties associated with sampling their habitats, we have limited information about which species live where, or what portion of their diversity remains to be discovered and described [@Mora2011]. Additionally, coordinated marine ecosystem monitoring initiatives have been lacking [@Duffy2013,@Borja2016] (but are now getting organized, see MarineGEO [@Duffy2016marinegeo], MBON, and PISCO), making it difficult to monitor changes in species distributions and abundances. While most documented global extinctions of marine species are vertebrates, many other species may have been lost unnoticed because they were not known [@Sala2006]. There is therefore a pressing need to improve our knowledge of marine systems before species go extinct, and their functioning is irreversibly altered.

Research attempting to understand how much we know about marine biodiversity has focused on estimating the total number of species that inhabits the oceans, and which fraction of this diversity has been named [@Appeltans2012,@Costello2013]. Undoubtedly, the development of molecular methods to investigate species limits has reinvigorated taxonomy and contributed to elucidating cryptic species complexes [@Knowlton2000]. Thus, it has been argued that molecular data can speed up the enterprise of discovering and naming species [@Blaxter2016]. As species remain the most convenient unit to measure biological diversity, understanding how much has been done, how much remains to be done, and how we can speed up the description of species diversity is crucial. However, to understand the role of these species in their environment, we need to document their ecology [@Cernansky2017]. For instance, to be able to predict how their geographic ranges will be affected by global change, we need detailed information about their environmental requirements. Occurrence data for marine species could be used for environmental niche modeling of their geographic and bathymetric distributions but the number of species -- and therefore the proportion of all species -- for which we have enough data to do so is unknown.

Until very recently, our knowledge of biodiversity was scattered among heterogeneous sources (e.g. natural history collections, publications, expert knowledge), difficult to access, and therefore difficult to synthesize. The development of public databases for specimen records (e.g. iDigBio [@Page2015]), species occurrences (e.g., OBIS [@Zhang2002]), genetic data (e.g., GenBank [@Benson2013], BOLD [@Ratnasingham2007a]), and registers of species names (e.g., WoRMS [@Costello2013b]) now allow us to evaluate what and how much we know. However, few studies have sought to synthesize existing knowledge by combining information from these databases.  Here, we aggregate publicly available data for marine invertebrates that occur within the limits of the exclusive economic zone (EEZ) of the continental United States (US), as well as the Gulf of Mexico (GOM) and the Pacific Northwest (PNW), to highlight gaps in our knowledge that provide a road map for important new directions in global research. We show that while public databases are capturing a significant portion of our knowledge about species richness, most species are known from only a few specimens and a few locations. We emphasize that our understanding of marine biodiversity, and the impact of human activities on biological communities require not only a more accurate list of the species that inhabit the oceans, but also more comprehensive data on species distributions and abundances. These goals can only be achieved with additional cost-effective sampling and a commitment to training more taxonomists and experts that can identify and interpret the material collected.

We focus on the continental US because the marine biodiversity is extensively documented as the result of a long history of studies by governmental, academic, and private institutions [@Fautin2010]. The continental US includes a wide diversity of marine ecosystems from the Florida Keys, to the Gulf of Maine on the East Coast; and from the arid Southern California to the oceanic shores of the Pacific Northwest on the West coast. With estuaries, bays, sandy beaches, rocky shores, and coral reefs, the US coastlines harbors many biota. While the continental shelf extends for several hundred kilometers on the East coast, it is narrow on the West coast. This diversity of habitats and bathymetric profiles, combined with efforts at the national level to make data publicly available, make this region especially suitable to evaluate how much we know about marine biodiversity using biodiversity repositories. Additionally, apart from the syntheses that emerged from the Census of Marine Life efforts [@Costello2010,@Fautin2010], no study has yet attempted to synthesize the current state of knowledge of marine biodiversity across the continental US, and this study is the first using data from public databases.

@@

>>>>

---
title: "The state of knowledge for marine invertebrate biodiversity in the continental US"
date: ' ' #'`r format(Sys.time())`'
output:
  bookdown::pdf_document2:
    includes:
    #  before_body: ../template/doc_prefix.tex
      in_header: preamble.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc: no
fontsize: 12pt
link-citations: yes
csl: nature.csl
bibliography: /home/francois/Library/dark-plankton.bib
abstract: Museum collections have been recording our knowledge about the diversity and distribution of species for about 250 years. Until recently, this information was fragmented across institutions and difficult to access or synthesize. The digitization of biodiversity information is enabling the synthesis of collection data. Here we use publicly accessible digital biodiversity data to asses how well we know the marine invertebrate fauna of the continental United States. Globally, marine inverterbrates comprise about 75% of described marine species. We document `r format(n_distinct(combined_species$worms_valid_name), big.mark = ",")` species from the continental US waters in digital databases (OBIS and iDigBio), and explore spatial and temporal patterns of their distribution and collection. Databases do well at the fauna level, capturing ~xxx% of the species diversity. However, sampling bias obscure broad diversity patterns, such as differences between coasts and across latitude. Our knowledge about species richness and geographic range is limited by sampling effort, with `r prop_10_specimens_species`% of species known from $\leq$10 samples. While recent developments in high-throughput sequencing promise to greatly speed up documentation of marine biodiversity, interpretation of sequences requires anchoring results anchoring results to identified vouchers. DNA barcode sequences (COI) are presently available for only `r format_output(prop_spp_barcoded$p_barcoded * 100)`% of the US marine invertebrates, and many of these are from samples taken outside the US. Over a third of the barcoded species are complexes of of multiple cryptic lineages, indicating that the fauna is substantially richer than the above number indicate and needs to be taxonomically reevaluated as a whole. Substantially increased sampling, coordinated at the national scale, is needed to provide a meaningful baseline for marine biodiversity and for metabarcoding and metagenomics studies. This data will allow the detection of the impacts of climate change and other anthropogenic strains on marine biodiversity. Advances in computing, information systems, and sequencing technologies, have set the stage for a broad understanding of the dynamics of marine biodiversity, but this potentially is limited by the availablility of current samples, digitization, and taxonomic effort.
---

@@

With increasing impact of human activities on the biosphere, we need efficient and accurate approaches to document changes in species diversity, distribution, and community composition. <!-- add why? -->  While terrestrial ecosystems have received considerable attention, the baseline of healthy and resilient marine ecosystems are not well understood [@Worm2006], and modeling the impact of human activities has been limited to a few well-studied taxonomic groups [@Tittensor2010]. Marine biodiversity remains poorly understood, with even the global richness of macrobiota poorly constrained[@Mora2011,@Appeltans2012,@Poore2015], and most species are known from few records and studies [@Bouchet2002]. Constrained by access to habitats and limited biodiversity infrastructure, coordinated, broad marine ecosyste monitoring initiatives have been slow to develop [@Duffy2013,@Borja2016] (but see for instance MarineGEO [@Duffy2016marinegeo], MBON, and PISCO), making it difficult to monitor changes in species distributions and community composition. Climate change and other human impacts are leading to rapid changes in marine ecosystems, and there is a pressing need to improve our knowledge of baseline marine systems and species diversity before they are further altered.

Biodiversity is fundamental to ecosystem function, but documenting and tracking biodiversity is laden with taxonomic and sampling challenges. For macrobiota, species occurrence has generally been documented through visual identification of live or captured specimens. Reliably-identified samples can then be used to generate molecular identification tools, such as DNA barcodes, opening additional methods for species documentation.

Species occurrence data is the basis for understanding species diversity and distributions, and is essential for monitoring community composition and ecosystem function. When species distributions are sufficiently well-sampled, occurrence data can also be used for environmental niche modelling to predict geographic and bathymetric distributions, community composition, and response of species and communities to changing environmental conditions. In well-sampled systems, like land plants, these can yield detailed prediction of future ecosystems (xxxx).

Until recently, our knowledge of biodiversity was scattered among heterogeneous sources (e.g., natural history collections, publications, expert knowledge) that are difficult to access and synthesize. The development of public databases for specimen records (e.g., iDigBio [@Page2015]), species occurrences (e.g., OBIS [@Zhang2002]), genetic data (e.g., GenBank [@Benson2013], BOLD [@Ratnasingham2007a]), and registers of species names (e.g., WoRMS [@Costello2013b]) have revolutionized how biodiversity research by allowing ready access to aggregated data to answer broad questions. However, few studies have sought to synthesize existing knowledge by combining information from these databases.

The goal of this study is to evaluate the extent of, and gaps in, our digital knowledge of species occurrences of marine invertebrates in the marine waters of the continental United States, one of the best documented areas in the world ocean. We aggregate publicly available data within the limits of the exclusive economic zone (EEZ) of the US, as well as the Gulf of Mexico (GOM) and the Pacific Northwest (PNW). We show that while public databases are capturing a significant portion of our knowledge about species richness, most species are known from only a few specimens and locations. We emphasize that our understanding of marine biodiversity, and the impact of human activities on biological communities require not only a more accurate knowledge of the species that inhabit the oceans, but also more comprehensive data on species distributions and abundances. These goals can only be achieved with additional cost-effective sampling and a commitment to developing and maintaining the taxonomic expertise who can identify and interpret the material collected.

We focus on the continental US because the marine biodiversity is extensively documented as a result of a long history of studies by governmental, academic, and private institutions [@Fautin2010]. The continental US includes a wide range of marine ecosystems from the Gulf of Mexico and Florida Keys, to the Gulf of Maine on the East Coast; and from arid Southern California to the shores of the Pacific Northwest on the West coast. With estuaries, bays, sandy beaches, rocky shores, and coral reefs, the US coastlines harbors many biota. While the continental shelf extends for several hundred kilometers on the East coast, it is narrow on the West coast. This diversity of habitats and bathymetric profiles, combined with efforts at the national level to make data publicly available, make this region especially suitable to evaluate how much we know about marine biodiversity using biodiversity repositories. Additionally, apart from the syntheses that emerged from the Census of Marine Life efforts [@Costello2010,@Fautin2010], no study has yet attempted to synthesize the current state of knowledge of marine biodiversity across the continental US, and this study is the first using data from public databases.

@@

<!---- Research attempting to understand how much we know about marine biodiversity has focused on estimating the total number of species that inhabits the oceans, and which fraction of this diversity has been named [@Appeltans2012,@Costello2013]. Undoubtedly, the development of molecular methods to investigate species limits has reinvigorated taxonomy and contributed to elucidating cryptic species complexes [@Knowlton2000]. Thus, it has been argued that molecular data can speed up the enterprise of discovering and naming species [@Blaxter2016]. As species remain the most convenient unit to measure biological diversity, understanding how much has been done, how much remains to be done, and how we can speed up the description of species diversity is crucial. However, to understand the role of these species in their environment, we need to document their ecology [@Cernansky2017]. For instance, to be able to predict how their geographic ranges will be affected by global change, we need detailed information about their environmental requirements. Occurrence data for marine species could be used for environmental niche modeling of their geographic and bathymetric distributions but the number of species -- and therefore the proportion of all species -- for which we have enough data to do so is unknown. -->
